---
title: "example workflow with tidypopgen"
output: rmarkdown::html_vignette
        #pdf_document
vignette: >
  %\VignetteIndexEntry{example_workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Ideally redo this with a dataset that does have some intermediates (e.g. subset from the HGDP)

## An example workflow with real data
We will explore the genetic structure of *Anolis punctatus* in South America,
using data from Prates et al 2018. We downloaded the vcf file of the genotypes
from "https://github.com/ivanprates/2018_Anolis_EcolEvol/blob/master/data/VCFtools_SNMF_punctatus_t70_s10_n46/punctatus_t70_s10_n46_filtered.recode.vcf?raw=true" and compressed it to a vcf.gz file.

We read in the data from the compressed vcf with:
```{r}
library(tidypopgen)
vcf_path <- system.file("/extdata/anolis/punctatus_t70_s10_n46_filtered.recode.vcf.gz",
                          package = "tidypopgen")
bed_path <- gt_vcf_to_bed(vcf_path, bed_path = tempfile("anolis_"))
anole_gt <- gen_tibble(bed_path)
```
Note that the `gt_vcf_to_bed` function reads the whole vcf in memory before writing
the bed, so it is only suitable for smallish datasets.

Now let's inspect our `gen_tibble`:
```{r}
anole_gt
```

We can see that we have 46 individuals, from 3249 loci. Note that we don't have any
information on population from the vcf. That information can be found
from another file on the github repository ("https://github.com/ivanprates/2018_Anolis_EcolEvol/raw/master/data/plot_order_punctatus_n46.csv).
We will have add the population information manually. Let's start by reading the
file:
```{r}
pops_path <- system.file("/extdata/anolis/plot_order_punctatus_n46.csv",
                        package = "tidypopgen")
pops <- readr::read_csv(pops_path)
pops
```

The ids from the VCF are in a different format than the ones we just got from
the pop csv. We need a bit of string wrangling, but it looks easy, we just need
to remove "punc_":

Let us simplify the ids, which wll have a "punc_" prefix
```{r}
anole_gt <- anole_gt %>% mutate(id = gsub('punc_',"",.data$id,))
anole_gt
```

Now we can bring in the pop information:
```{r}
anole_gt <- anole_gt %>% mutate(population = pops$pop[match(pops$ID,.data$id)])
anole_gt
```

That was easy. The loci had already been filtered and cleaned, so we don't need 
to do any QC. Let us jump straight into analysis and run a PCA:
```{r, error=TRUE}
anole_pca <- anole_gt %>% gt_pca_partialSVD(k=10)
```
Ok, we jumped too quickly. There are missing data, and we need first 
to impute them:

```{r}
anole_gt <- gt_impute_simple(anole_gt)
```

And now:
```{r}
anole_pca <- anole_gt %>% gt_pca_partialSVD(k=45)
```



Let us add the PC scores to our `gen_tibble`:
```{r}
anole_gt <- augment(anole_pca , data = anole_gt)
library(ggplot2)
anole_gt %>% ggplot(aes(.fittedPC1, .fittedPC2, color = population)) + 
  geom_point(size = 1.5)
```
We can see that the three population do separate nicely on the PCA.

# Explore population structure with DAPC

DAPC is a powerful tool to investigate population structure. It has the advantage of
scaling well to very large datasets, and it does not have the assumptions of STRUCTURE or
ADMIXTURE (but that can also be a drawback).

The first step is to determine the number of genetic clusters in the dataset. We will use
a k-clustering algorithm applied to the principal components (allowing us to reduce
the dimensions from the thousands of loci to just a few tens components). We need to decide how 
many components to use; this decision is often made based on a plot of the cumulative
explained variance of the components.
Using `tidy` on the `gt_pca` object allows us easily obtain those quantities, and it is then
trivial to plot them:
```{r}
library(ggplot2)
tidy(anole_pca,matrix="eigenvalues") %>%
  ggplot(mapping =aes(x=PC, y=cumulative)) +
  geom_point()

```

Ideally, we are looking for the point where the curve starts flattening. In this case, we can
not see a very clear flattening, but by PC 30 the increase in explained variance as markedly
decelerated. As the k-clustering algorithm is somewhat data hungry,
we will go with that somewhat generous number

```{r}
anole_clusters <- gt_pca_find_clusters(anole_pca, n_pca = 40)
```

This object retains information about a large number of possible *k*. We can choose
the best k with `gt_pca_clust_choose_n()` (we will use the defaults, but there
are multiple statistics and criteria that we can use to choose *n*):
```{r}
anole_clusters <- gt_pca_best_k(anole_clusters)
```



Now we can run a DAPC:
```{r}
anole_dapc <- gt_dapc(anole_clusters)
```

Note that `gt_dapc()` takes automatically 3 clusters form the `anole_clusters` object,
but can change that behaviour by setting some of its parameters (see the help
page for `gt_dapc()`).
`anole_dapc` is of class `dapc`, so we can use the standard `adegenet` functions:
```{r}
anole_dapc
scatter(anole_dapc)
```


```{r}
dapc_data_df <-
  # as_tibble() converts the ind.coord matrix to a special data frame called a tibble. 
  as_tibble(anole_dapc$ind.coord, rownames = "individual") %>%
  # mutate changes or adds columns to a data frame. Here, we're adding the population and group assignment columns to the data frame
  mutate(population = pops$pop,
         group = anole_dapc$grp)
dapc_plot <-
  ggplot(dapc_data_df, aes(
    x = LD1,
    y = LD2,
    fill = population
  )) +
  geom_point(shape = 21, size = 3) +
  #reverse the color direction to better reflect Prates et al. 2018
  scale_fill_viridis_d(direction = -1) + 
  theme_bw(base_size = 16)

dapc_plot

```





We can inspect the assignment by DAPC with `assignplot`, ordering the samples
by the original population label:
```{r}
assignplot(anole_dapc,subset=order(anole_gt$population))
```


As one would expect from the clear separation in the PCA and DAPC scatterplots,
individuals are all fully assigned to a single cluster, with no individual with
mixed ancestry.

```{r}
compoplot(anole_dapc, subset=order(anole_gt$population), posi="bottomright",
txt.leg=paste("Cluster", 1:5), lab="",
ncol=2, xlab="individuals", col=funky(5))
```
