---
title: "Example workflow with tidypopgen"
output: rmarkdown::html_vignette
        #pdf_document
vignette: >
  %\VignetteIndexEntry{Example workflow with tidypopgen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
```


## An example workflow with real data
We will explore the genetic structure of *Anolis punctatus* in South America,
using data from Prates et al 2018. We downloaded the vcf file of the genotypes
from "https://github.com/ivanprates/2018_Anolis_EcolEvol/blob/master/data/VCFtools_SNMF_punctatus_t70_s10_n46/punctatus_t70_s10_n46_filtered.recode.vcf?raw=true" and compressed it to a vcf.gz file.

We read in the data from the compressed vcf with:
```{r}
library(tidypopgen)
vcf_path <- system.file("/extdata/anolis/punctatus_t70_s10_n46_filtered.recode.vcf.gz",
                          package = "tidypopgen")
anole_gt <- gen_tibble(vcf_path, quiet = TRUE, backingfile = tempfile("anolis_"))

```

Now let's inspect our `gen_tibble`:
```{r}
anole_gt
```

We can see that we have 46 individuals, from 3249 loci. Note that we don't have any
information on population from the vcf. That information can be found
from another file on the github repository ("https://github.com/ivanprates/2018_Anolis_EcolEvol/raw/master/data/plot_order_punctatus_n46.csv).
We will have add the population information manually. Let's start by reading the
file:
```{r}
pops_path <- system.file("/extdata/anolis/plot_order_punctatus_n46.csv",
                        package = "tidypopgen")
pops <- readr::read_csv(pops_path)
pops
```

The ids from the VCF are in a different format than the ones we just got from
the pop csv. We need a bit of string wrangling, but it looks easy, we just need
to remove "punc_":

Let us simplify the ids, which will have a "punc_" prefix
```{r}
anole_gt <- anole_gt %>% mutate(id = gsub('punc_',"",.data$id,))
anole_gt
```

Now we can bring in the pop information:
```{r}
anole_gt <- anole_gt %>% mutate(population = pops$pop[match(pops$ID,.data$id)])
anole_gt
```

## PCA

That was easy. The loci had already been filtered and cleaned, so we don't need 
to do any QC. Let us jump straight into analysis and run a PCA:
```{r, error=TRUE}
anole_pca <- anole_gt %>% gt_pca_partialSVD(k=30)
```
OK, we jumped too quickly. There are missing data, and we need first 
to impute them:

```{r}
anole_gt <- gt_impute_simple(anole_gt, method = "mode")
```

And now:
```{r}
anole_pca <- anole_gt %>% gt_pca_partialSVD(k=30)
```

Let us look at the object:
```{r}
anole_pca
```

The `print` function (implicitly called when we type the name of the object)
gives us information about the most important elements in the object (and the names
of the elements in which they are stored).

We can extract those elements with the `tidy` function, which returns a tibble
that can be easily used for further analysis, e.g.:
```{r}
tidy(anole_pca, matrix="eigenvalues")

```
We can return information on the *eigenvalues*, *scores* and *loadings* of the pca.
There is also an `autoplot` method that allows to visualise those elements (type *screeplot* for
*eigenvalues*, type *scores* for *scores*, and *loadings* for *loadings*:
```{r}
autoplot(anole_pca, type="screeplot")
```

To plot the sample in principal coordinates space, we can simply use:
```{r}
autoplot(anole_pca, type ="scores")
```

`autoplots` are deliberately kept simple: they are just a way to quickly inspect the results.
They generate `ggplot2` objects, and so they can be further embellished with the usual
`ggplot2` grammar:
```{r}
library(ggplot2)
autoplot(anole_pca, type = "scores") +
  aes(color = anole_gt$population) +
  labs(color = "population")

```

For more complex/publication ready plots, we will want to add the PC scores
to the tibble, so that we can create a custom plot with `ggplot2`. We can easily
add the data with the `augment` method:

```{r}
anole_gt <- augment(anole_pca , data = anole_gt)
```

And now we can use `ggplot2` directly to generate our plot:

```{r}
anole_gt %>% ggplot(aes(.fittedPC1, .fittedPC2, color = population)) + 
  geom_point()
```

We can see that the three population do separate nicely on the PCA, with just one individual
from Wam sitting in-between the other Wam individuals and those from Eam.

It is also possible to inspect which loci contribute the most to a given component:

```{r}
autoplot(anole_pca, type = "loadings")
```

By using information from the loci table, we could easily embellish the plot, for example
colouring by chromosome or maf.

For more complex plots, we can augment the loci table with the loadings using `augment_loci()`:

```{r}
anole_gt_load <- augment_loci(anole_pca, data= anole_gt)

```

# Explore population structure with DAPC

DAPC is a powerful tool to investigate population structure. It has the advantage of
scaling well to very large datasets. It does not have the assumptions of STRUCTURE or
ADMIXTURE (which also limits its power).

The first step is to determine the number of genetic clusters in the dataset. 
DAPC can be either used to test a a-priori hypothesis, or we can use the data to
suggest the number of clusters.
In this
case, we did not have any strong expectations of structure in our study system, so we will
let the data inform the number of possible genetic clusters. We will use
a k-clustering algorithm applied to the principal components (allowing us to reduce
the dimensions from the thousands of loci to just a few tens of components). We need to decide how 
many components to use; this decision is often made based on a plot of the cumulative
explained variance of the components.
Using `tidy` on the `gt_pca` object allows us easily obtain those quantities, and it is then
trivial to plot them:
```{r}
library(ggplot2)
tidy(anole_pca,matrix="eigenvalues") %>%
  ggplot(mapping =aes(x=PC, y=cumulative)) +
  geom_point()

```

Note that, as we were working with a truncated SVD algorithm for our PCA, we can not easily
phrase the eigenvalues in terms of proportion of total variance, so the cumulative y axis simply
shows the cumulative sum of the eigenvalues. Ideally, we are looking for the point where 
the curve starts flattening. In this case, we can
not see a very clear flattening, but by PC 10 the increase in explained variance has markedly
decelerated. We can now find clusters based on those 10 PCs:

```{r}
anole_clusters <- gt_cluster_pca(anole_pca, n_pca = 10)
```

As we did not define the *k* values to explore, the default 1 to 5 was used (we
can change that by setting the `k` parameter to change the range). To choose an appropriate
*k*, we plot the number of clusters against a measure of fit. BIC has been shown
to be a very good metric under many scenarios:
```{r}
autoplot(anole_clusters)
```

We are looking for the minimum value of BIC. There is no clear elbow (a minimum after which BIC increases
with increasing k). However, we notice that there is a quick levelling
off in the decrease in BIC at
3 clusters. Arguably, these are sufficient to capture the main structure (and that makes
sense given what we saw in the PCA). We can also use a number of algorithmic approaches (based on the original `find.clusters()` function in `adegenet`) to choose the best *k* value
from this plot through `gt_cluster_pca_best_k()`. We will use the defaults (BIC with "diffNgroup",
see the help page for `gt_cluster_pca_best_k()` for a description of the various options):
```{r}
anole_clusters <- gt_cluster_pca_best_k(anole_clusters)
```
The algorithm confirms our choice. Note that this function simply adds an element
`$best_k` to the `gt_cluster_pca` object:
```{r}
anole_clusters$best_k
```
If we decided that we wanted to explore a different value, we could simply overwrite
that number with `anole_clusters$best_k<-5`

In this case, we are happy with the option of 3 clusters, and we can run a DAPC:
```{r}
anole_dapc <- gt_dapc(anole_clusters)
```

Note that `gt_dapc()` takes automatically the number of clusters from the `anole_clusters` object,
but can change that behaviour by setting some of its parameters (see the help
page for `gt_dapc()`). When we print the object, we are given information about
the most important elements of the object and where to find them (as we saw for `gt_pca`):

```{r}
anole_dapc
```

Again, these elements can be obtained with tidiers (with `matrix` equal to `eigenvalues`,
`scores`,`ld_loadings` and `loci_loadings`):
```{r}
tidy(anole_dapc, matrix="eigenvalues")
```

And they can be visualised with `autoplot`:
```{r}
autoplot(anole_dapc, type="screeplot")
```

As for pca, there is a `tidy` method that can be used to extract information
from `gt_dapc` objects. For example, if we want to create a bar plot of
the eigenvalues (since we only have two), we could simply use:
```{r}
tidy(anole_dapc, matrix="eigenvalues") %>%
  ggplot(aes(x=LD,y=eigenvalue)) +
  geom_col()
```

We can plot the scores with:
```{r plot_scatter_dapc}
autoplot(anole_dapc, type="scores")
```

We can inspect the assignment by DAPC with `autoplot` using the type
`components`, ordering the samples
by the original population labels:

```{r}
autoplot(anole_dapc, type="components", group = anole_gt$population)
```

Because of the very clear separation we observed when plotting the LD scores,
no individual is modelled as a mixture: all assignments are with 100% probability 
to a single cluster.

Finally, we can explore which loci have the biggest impact on separating the clusters
(either because of drift or selection):

```{r}
autoplot(anole_dapc, "loadings")
```

There is no strong outlier, suggesting drift across many loci has created the signal
picked up by DAPC.

Note that `anole_dapc` is of class `gt_dapc`, which is a subclass of `dapc` 
from `adegenet`. This means that functions written to work on `dapc` objects
should work out of the box (the only exception is `adegenet::predict.dapc`, 
which does not work because the underlying pca object is different). For example,
we can obtain the standard dapc plot with:
```{r}
library(adegenet)
scatter(anole_dapc, posi.da="bottomright")
```



# Clustering with sNMF

sNMF is a fast clustering algorithm which provides results similar to STRUCTURE and ADMIXTURE. We can run it directly from R. We have first to generate a file
with our genotype data:
```{r}
geno_file <- gt_as_geno_lea(anole_gt)
geno_file
```

Note that the .geno file is placed by default in the same directory and using
the same name as the backing file of the `gen_tibble`

Now we can run K clusters from k=1 to k=10. We will use just one repeat, but ideally
we should run multiple repetition for each K:
```{r results='hide'}
library(LEA)
anole_snmf <- snmf(input.file = geno_file,
                   K = 1:10,
                   entropy = TRUE,
                   repetitions = 3,
                   alpha = 100
                  )
```

We can examine the suitability of our K values by plotting with: 

```{r}
plot(anole_snmf, cex = 1.2, pch = 19)
```

From which we can see that K = 3 is a sensible choice, as 3 represents the 'elbow' in the plot. We can select the fastest run using: 


```{r}
ce <-  cross.entropy(anole_snmf, K = 3)
quick_run <- which.min(ce)
```

And then to plot our clusters, we can begin by extracting the Q matrix from our `snmf` object,
and covert it to a `q_matrix` object as used by `tidypopgen`:

```{r}
q_mat <- LEA::Q(anole_snmf, K = 3, run = quick_run) 
q_mat <- as_q_matrix(q_mat)

head(q_mat)
```

We can quickly plot it with
```{r}
autoplot(q_mat,gen_tbl = anole_gt)
```

We can tidy our q matrix, and attach the population and id data from our original gen_tibble, returning it
in a format which is suitable for plotting:

```{r}
tidy_q <- tidy(q_mat, anole_gt)
head(tidy_q)
```
Note that we also add the dominant `q` for a given individual, as well as the dominant population which is associated
with that `q`

And we can plot the tibble directly with `ggplot`:

```{r}
# set up the ggplot object
plt <- ggplot2::ggplot(tidy_q,
                       ggplot2::aes(x = id,
                                    y = percentage,
                                    fill = q)) +
  # add the columns based on percentage membership to each cluster
  ggplot2::geom_col(width = 1,
                    position = ggplot2::position_stack(reverse = TRUE))+
  # set the y label
  ggplot2::labs(y = "K = 3")+
  # use a theme to match the distruct look, removing most decorations
  theme_distruct() +
  # set the colour scale to be the same as in distruct and clumppling
  scale_fill_distruct()

plt
```



